Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-03 paper/pool_model.tex
write next two sections of introduction

\noindent
\paragraph{Experimental/Biological evidence/phenomena}
\begin{itemize}
    \item Lag Phase, Competition, Cooperation, Limiting Resource
    \item Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}
    \item Spatial Effects: adhesion, migration, chemotaxis, diffusion
    \item Quorum-sensing
    \item Biofilm formation \cite{Jin2020}
\end{itemize}

\noindent
\paragraph{ODE models (predictive microbiology, existing, overview, most popular)}
Existing models in Predictive Microbiology (Primary Models)~\cite{perez-rodriguez_predictive_2012}:
\begin{itemize}
    \item An overview of primary models used in food science can be found in~\cite{van_boekel_kinetic_2008}.
    \item Baranyi-Roberts Model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994}
    \item modified Gompertz~\cite{Gompertz1825-wi,Gibson1987}
    \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\cite{buchanan_when_1997}
    \item McKellar two-compartment model with bacteria in growing and non-growing state~\cite{mckellar_heterogeneous_1997}
    \item Inactivation models
\end{itemize}
%
Secondary models:
\begin{itemize}
    \item Polynomial models 
    \item Square Root-Type Models, e.g., Ratkowski for temperature~\cite{ratkowsky_relationship_1982}
    \item The Cardinal Parameter Model~\cite{zwietering_decision_1992}
\end{itemize}

>>>>

% \noindent
% \paragraph{Experimental/Biological evidence/phenomena}
% \begin{itemize}
%     \item Lag Phase, Competition, Cooperation, Limiting Resource
%     \item Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}
%     \item Spatial Effects: adhesion, migration, chemotaxis, diffusion
%     \item Quorum-sensing
%     \item Biofilm formation \cite{Jin2020}
% \end{itemize}

\noindent
In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biologial phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabilic byproduct.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony.
Bacteria have thus devised various survival strategies to ensure a sufficient access to nutrients.
Effects on a single-cell level such as adhesion, migration and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial community.

% \noindent
% \paragraph{ODE models (predictive microbiology, existing, overview, most popular)}
% Existing models in Predictive Microbiology (Primary Models)~\cite{perez-rodriguez_predictive_2012}:
% \begin{itemize}
%     \item An overview of primary models used in food science can be found in~\cite{van_boekel_kinetic_2008}.
%     \item Baranyi-Roberts Model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994}
%     \item modified Gompertz~\cite{Gompertz1825-wi,Gibson1987}
%     \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\cite{buchanan_when_1997}
%     \item McKellar two-compartment model with bacteria in growing and non-growing state~\cite{mckellar_heterogeneous_1997}
%     \item Inactivation models
% \end{itemize}
% %
% Secondary models:
% \begin{itemize}
%     \item Polynomial models 
%     \item Square Root-Type Models, e.g., Ratkowski for temperature~\cite{ratkowsky_relationship_1982}
%     \item The Cardinal Parameter Model~\cite{zwietering_decision_1992}
% \end{itemize}

\noindent
The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under contant
environmental conditions.
Widely used examples include the modified Gompertz model~\cite{Gompertz1825-wi}, Baranyi-Roberts
model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994} and MCKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\cite{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\cite{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicitiy of usage.
However due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.
